---
title: "Getting Started with exnexSurv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with exnexSurv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Overview

`exnexSurv` fits Bayesian EXNEX survival models for right-censored log-normal data.
The current interface supports:

- a formula interface,
- an `x`/`y` interface,
- subgroup effects,
- optional covariates,
- posterior draws for `theta_j`, `beta`, and `sigma2`.

## Simulate a simple dataset

```{r}
library(exnexSurv)
library(survival)

simulate_surv_data <- function(
  theta,
  sigma2,
  beta = NULL,
  n_per_group = 50,
  censor_min = 4,
  censor_max = 12,
  seed = NULL
) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  groups <- rep(seq_along(theta), each = n_per_group)
  age_std <- rnorm(length(groups), mean = 0, sd = 1)
  mean_log_time <- rep(theta, each = n_per_group)

  if (!is.null(beta)) {
    mean_log_time <- mean_log_time + beta * age_std
  }

  log_time <- rnorm(length(groups), mean = mean_log_time, sd = sqrt(sigma2))
  true_time <- exp(log_time)
  censor_time <- runif(length(groups), min = censor_min, max = censor_max)

  data.frame(
    time = pmin(true_time, censor_time),
    event = as.integer(true_time <= censor_time),
    group = factor(groups),
    age_std = age_std
  )
}

sim_data <- simulate_surv_data(
  theta = c(1.1, 1.6, 2.0),
  sigma2 = 0.25,
  beta = -0.30,
  n_per_group = 50,
  seed = 6421
)

head(sim_data)
mean(sim_data$event)
```

## Fit the model with a formula

```{r}
fit_formula <- exnex_surv(
  Surv(time, event) ~ group + age_std,
  data = sim_data,
  iter = 1200,
  warmup = 400,
  chains = 1,
  seed = 6421
)

print(fit_formula, show_trace = FALSE)
summary(fit_formula)
```

The fitted object stores the processed data, posterior draws, and MCMC settings.
You can inspect the model structure with the usual S3 methods.

## Run multiple chains in parallel

```{r}
fit_parallel <- exnex_surv(
  Surv(time, event) ~ group + age_std,
  data = sim_data,
  iter = 1200,
  warmup = 400,
  chains = 2,
  parallel_chains = 2,
  seed = 6421
)

print(fit_parallel, show_trace = FALSE)
plot(
  fit_parallel,
  parameters = c("theta_1", "theta_2", "theta_3", "beta_1", "sigma2"),
  ask = FALSE
)
```

When `parallel_chains` is greater than 1, the chains are evaluated concurrently by R and the traceplots show all chains in the same panel for each parameter.

If you do not want to use all available workers, set `parallel_chains` smaller than `chains`. For example, `chains = 4` and `parallel_chains = 2` runs four independent chains while evaluating two at a time.

The fitted object stores the combined post-warmup draws from all chains in one table. Because each chain contributes the same number of post-warmup samples, the total number of rows is `(iter - warmup) * chains`.

```{r}
str(fit_formula$data)
exnexSurv:::plot.exnex_surv(
  fit_formula,
  parameters = c("theta_1", "theta_2", "sigma2"),
  ask = FALSE
)
```

## Fit the model with `x` and `y`

```{r}
fit_xy <- exnex_surv(
  x = sim_data[c("group", "age_std")],
  y = Surv(sim_data$time, sim_data$event),
  iter = 1200,
  warmup = 400,
  chains = 1,
  seed = 6421
)

summary(fit_xy)
all.equal(fit_formula$draws, fit_xy$draws)
```

This is useful if predictors and outcomes are prepared separately.

## Fit a model without covariates

If the model only contains the subgroup variable, no `beta` terms are estimated.

```{r}
fit_no_cov <- exnex_surv(
  Surv(time, event) ~ group,
  data = sim_data,
  iter = 1200,
  warmup = 400,
  chains = 1,
  seed = 6421
)

summary(fit_no_cov)
exnexSurv:::plot.exnex_surv(fit_no_cov, ask = FALSE)
```

## Basic posterior summaries

```{r}
summary(fit_formula)
```

## Notes

The current implementation supports multiple chains, including R-level parallel execution through `parallel_chains`. For a more careful convergence check, fit more than one chain and compare the traceplots and posterior summaries across chains.
